---
title: "FTIR"
output: html_document
date: "2024-11-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# FTIR PROCESSING - STATISTICAL ANALYSIS AND PLOTS

```{r}
library(readr)
library(opusreader2) #Open OPUS files
library(prospectr) #IR spectrum processing
library(ggplot2) #Plots
library(dplyr)
library(tidyr)
library(tidyverse)
library(tibble)
library(FactoMineR) #PCA analysis 
library(factoextra) #Plot PCA results
library(pROC)
library(MASS) #LDA
library(caret) #Classification and Regression Training

setwd("~/Library/CloudStorage/OneDrive-UniversitatdeBarcelona/PhD_A.Sayol/2024.IR_Biotyper/FTIR_R_analysis/Scripts_paper_FTIR/")
```

## HEALTHY DONORS

### Data loading

SCFAs levels from healthy donors and IR data files (OPUS files; .0):

```{r}
#SCFAs normalized levels - healthy donors:
scfa_hd <- read_csv("SCFA_normalised_HD.csv", col_names = TRUE, col_types = NULL)

#Create a variable named "sample_id_hd" to use it later for averageing spectra
sample_id_hd <- scfa_hd$sample_id

#Load FTIR data
spc_list_healthy <- read_opus(dsn = "~/Library/CloudStorage/OneDrive-UniversitatdeBarcelona/PhD_A.Sayol/2024.IR_Biotyper/FTIR_R_analysis/OPUS_spectra_FTIR/Healthy_donors_IR/", data_only = FALSE, parallel = FALSE, progress_bar = FALSE)

#We create an element for the wavenumbers (we could have used any sample as the x axis is the same for all the samples)
x_wave_list <- spc_list_healthy[["CLA33F(D_1)_D2.0"]][["ab"]][["wavenumbers"]]
```

### Spectra processing

Each sample has 5 replicates (except 2 of them, that have 4 replicates --\> CGR32F D.5 and MGM31F D.2), so we calculate the mean spectrum of each sample, having a single averaged spectrum per sample.

```{r}
#SPECTRA LIST --> Healthy donors
ir_names_healthy <- names(spc_list_healthy)

ir_healthy <- data.frame()
for (i in seq_along(sample_id_hd)) { #Iterate over each sample to get their 5 replicates
  pattern <- paste0("^", gsub("\\(", "\\\\(", gsub("\\)", "\\\\)", sample_id_hd[i])))
  replicates <- ir_names_healthy[grep(pattern, ir_names_healthy)]
  if (length(replicates) != 0){
      selected_matrices <- lapply(replicates, function(element) {
      return(t(spc_list_healthy[[element]]$ab$data))
  })
  df_with_all_replicates <- as.data.frame(do.call(cbind, selected_matrices))
  df_for_each_sample <- data.frame(row.names = sample_id_hd[i],t((average_spec = rowMeans(df_with_all_replicates))))
  ir_healthy <- rbind(ir_healthy, df_for_each_sample)
  }
}

#Remove the X before each wavelenght
colnames(ir_healthy) <- gsub("^X", "", colnames(ir_healthy))

#Data sorting (raw IR spectra) 
raw_absorbance_hd <- ir_healthy %>% 
  rownames_to_column(var = "sample_id") %>%  #Row names to column
  pivot_longer(cols = -sample_id,            #Select all columns except "sample_id"
               names_to = "wavelength",     #Name for the wavelength column
               values_to = "absorbance")    #Name for the absorbance values

#Remove the 'X' before the wavelength value 
raw_absorbance_hd$wavelength <- gsub("X", "", raw_absorbance_hd$wavelength)  
raw_absorbance_hd$wavelength <- as.numeric(raw_absorbance_hd$wavelength)
```

**a) Standard normal variate (SNV)**

Facilitates interpretation as it passes the spectrum to whole numbers. Performed with the average spectra.

NOTE: The spectra shape does not change, but changes to whole numbers (absorbance), making it easier to understand.

**b) Baseline correction**

Appplied to SNV corrected data

**c) Savitzky-Golay**

We apply the second derivative to the curve. A Savitzky--Golay filter is a digital filter that can be applied to a set of digital data points for the purpose of smoothing the data, that is, to increase the precision of the data without distorting the digital tendency. The extremes of the dataframe are not included (the first and lasts wavelengths are omitted).

Parameters:

m = 2 (an integer indicating the differentiation order, derivative order)

p = 3 (an integer indicating the polynomial order)

w = 9 (window size, must be odd)

```{r}
#SPECTRA PROCESSING 
#A) STANDARD NORMAL VARIATE --> With average data (the spectra shape does not change, but changes to enter numbers, making it easier to understand)
ir_healthy_snv <- data.frame()
for (i in 1:nrow(ir_healthy)){
  print(i)
  ir_healthy_snv <- rbind(ir_healthy_snv,data.frame(standardNormalVariate(ir_healthy[i,])))
}
rownames(ir_healthy_snv) <- rownames(ir_healthy)
ir_healthy_snv

#Data sorting (STANDARD NORMAL VARIATE DATA)
snv_data_hd <- ir_healthy_snv %>% 
  rownames_to_column(var = "sample_id") %>%  #Row names to column
  pivot_longer(cols = -sample_id,            #Select all columns except "sample_id"
               names_to = "wavelength",     #Name for the wavelength column
               values_to = "absorbance")    #Name for the absorbance values

#Remove the 'X' before the wavelength value 
snv_data_hd$wavelength <- gsub("X", "", snv_data_hd$wavelength)  
snv_data_hd$wavelength <- as.numeric(snv_data_hd$wavelength)

#B) BASELINE CORRECTION --> Applied to SNV corrected data
ir_healthy_bsln <- data.frame()
for (i in 1:nrow(ir_healthy_snv)){
  numeric_vector <- ir_healthy_snv[i,]
  row_df <- data.frame(matrix(baseline(numeric_vector), nrow = 1, ncol = length(numeric_vector)))
  ir_healthy_bsln <- rbind(ir_healthy_bsln,row_df)
}

colnames(ir_healthy_bsln) <- colnames(ir_healthy)
rownames(ir_healthy_bsln) <- rownames(ir_healthy)

#Data sorting (BASELINE CORRECTED DATA)
bsln_data_hd <- ir_healthy_bsln %>% 
  rownames_to_column(var = "sample_id") %>%  #Row names to column
  pivot_longer(cols = -sample_id,            #Select all columns except "sample_id"
               names_to = "wavelength",     #Name for the wavelength column
               values_to = "absorbance")    #Name for the absorbance values

#Remove the 'X' before the wavelength value 
bsln_data_hd$wavelength <- gsub("X", "", bsln_data_hd$wavelength)  
bsln_data_hd$wavelength <- as.numeric(bsln_data_hd$wavelength)

#SNV + baseline correction plot (per sample) --> Check data visualization
blsn_hd_plot <- ggplot(bsln_data_hd, aes(x = wavelength, y = absorbance, color = sample_id)) + 
  geom_line(linewidth = 0.1) + 
  geom_hline(yintercept=0.0, linetype="dashed",color = "red", linewidth = 0.2) +
  labs(x = expression("Wavenumber (cm"^-1*")"), y = "Absorbance (a.u.)", title = "SNV + baseline corrected spectra - Healthy donors") + 
  scale_x_continuous(breaks = seq(min(bsln_data_hd$wavelength), max(bsln_data_hd$wavelength), by = 200), labels = seq(min(bsln_data_hd$wavelength), max(bsln_data_hd$wavelength), by = 200), expand = c(0, 0), trans = "reverse") +
  theme_minimal() +
  theme(text = element_text(size = 10),
        axis.title.y = element_text(vjust = +3, size = 12), 
        axis.title.x = element_text(vjust = +1, size = 12), 
        axis.ticks.x=element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none", 
        legend.text=element_text(size = 10))

blsn_hd_plot

#C) SAVITZKY-GOLAY --> 2nd derivate
ir_healthy_skg <- data.frame()
for (i in 1:nrow(ir_healthy_bsln)){
  numeric_vector <- ir_healthy_bsln[i,]
  row_df <- data.frame(savitzkyGolay(X = numeric_vector, m = 2, p = 3, w = 9))
  ir_healthy_skg <- rbind(ir_healthy_skg,row_df)
}

#Remove the X before each wavelenght
colnames(ir_healthy_skg) <- gsub("^X", "", colnames(ir_healthy_skg))

#Data sorting (SAVITZKY-GOLAY CORRECTED DATA)
skg_data_hd <- ir_healthy_skg %>% 
  rownames_to_column(var = "sample_id") %>%  #Row names to column
  pivot_longer(cols = -sample_id,            #Select all columns except "sample_id"
               names_to = "wavelength",     #Name for the wavelength column
               values_to = "absorbance")    #Name for the absorbance values

#Remove the 'X' before the wavelength value 
skg_data_hd$wavelength <- gsub("X", "", skg_data_hd$wavelength)  
skg_data_hd$wavelength <- as.numeric(skg_data_hd$wavelength)
```

## CDI PATIENTS

### Data loading

SCFAs levels from patients with *Clostidioides difficile* infection (CDI) and IR data files (OPUS files; .0):

```{r}
#Raw data - C.difficile patients:
scfa_cdiff <- read_csv("SCFA_normalised_CDIFF.csv", col_names = TRUE, col_types = NULL)

#Create a variable named "sample_id_hd" to use it later for averageing spectra
sample_id_cdiff <- scfa_cdiff$sample_id

#Load FTIR data
spc_list_cdiff <- read_opus(dsn = "~/Library/CloudStorage/OneDrive-UniversitatdeBarcelona/PhD_A.Sayol/2024.IR_Biotyper/FTIR_R_analysis/OPUS_spectra_FTIR/Cdifficile_IR/", data_only = FALSE, parallel = FALSE, progress_bar = FALSE)
```

### Spectra processing

Each sample has 5 replicates, so we calculate the mean spectrum of each sample, having a single averaged spectrum per sample.

```{r}
#SPECTRA LIST --> C. difficile pateints
ir_names_cdiff <- names(spc_list_cdiff)

ir_cdiff <- data.frame()
for (i in seq_along(sample_id_cdiff)) { # iterate over each sample to get their 5 replicates
  replicates <- ir_names_cdiff[grep(sample_id_cdiff[i], ir_names_cdiff)]
  if (length(replicates) != 0){
      selected_matrices <- lapply(replicates, function(element) {
      return(t(spc_list_cdiff[[element]]$ab$data))
  })
  df_with_all_replicates <- as.data.frame(do.call(cbind, selected_matrices))
  df_for_each_sample <- data.frame(row.names = sample_id_cdiff[i],t((average_spec = rowMeans(df_with_all_replicates))))
  ir_cdiff <- rbind(ir_cdiff, df_for_each_sample)
  }
}

#Remove the X before each wavelenght
colnames(ir_cdiff) <- gsub("^X", "", colnames(ir_cdiff))

#Data sorting (raw IR spectra) 
raw_absorbance_cdiff <- ir_cdiff %>% 
  rownames_to_column(var = "sample_id") %>%  #Row names to column
  pivot_longer(cols = -sample_id,            #Select all columns except "sample_id"
               names_to = "wavelength",     #Name for the wavelength column
               values_to = "absorbance")    #Name for the absorbance values

#Remove the 'X' before the wavelength value 
raw_absorbance_cdiff$wavelength <- gsub("X", "", raw_absorbance_cdiff$wavelength)  
raw_absorbance_cdiff$wavelength <- as.numeric(raw_absorbance_cdiff$wavelength)
```

Processing:

```{r}
#SPECTRA PROCESSING
#A) STANDARD NORMAL VARIATE --> With average data (the spectra form does not change, but changes to enter numbers, making it easier to understand)
ir_cdiff_snv <- data.frame()
for (i in 1:nrow(ir_cdiff)){
  print(i)
  ir_cdiff_snv <- rbind(ir_cdiff_snv,data.frame(standardNormalVariate(ir_cdiff[i,])))
}
rownames(ir_cdiff_snv) <- rownames(ir_cdiff)
ir_cdiff_snv

#Data sorting (STANDARD NORMAL VARIATE DATA)
snv_data_cdiff <- ir_cdiff_snv %>% 
  rownames_to_column(var = "sample_id") %>%  #Row names to column
  pivot_longer(cols = -sample_id,            #Select all columns except "sample_id"
               names_to = "wavelength",     #Name for the wavelength column
               values_to = "absorbance")    #Name for the absorbance values

#Remove the 'X' before the wavelength value 
snv_data_cdiff$wavelength <- gsub("X", "", snv_data_cdiff$wavelength)  
snv_data_cdiff$wavelength <- as.numeric(snv_data_cdiff$wavelength)

#SNV plot (per sample) --> Check data visualization
snv_cdiff_plot <- ggplot(snv_data_cdiff, aes(x = wavelength, y = absorbance, color = sample_id)) + 
  geom_line(linewidth = 0.1) + 
  geom_hline(yintercept=0.0, linetype="dashed",color = "red", linewidth = 0.2) +
  labs(x = expression("Wavenumber (cm"^-1*")"), y = "Absorbance (a.u.)", title = "SNV corrected spectra - C.difficile patients") + 
  scale_x_continuous(breaks = seq(min(snv_data_cdiff$wavelength), max(snv_data_cdiff$wavelength), by = 200), labels = seq(min(snv_data_cdiff$wavelength), max(snv_data_cdiff$wavelength), by = 200), expand = c(0, 0), trans = "reverse") +
  theme_minimal() +
  theme(text = element_text(size = 10),
        axis.title.y = element_text(vjust = +3, size = 12), 
        axis.title.x = element_text(vjust = +1, size = 12), 
        axis.ticks.x=element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none", 
        legend.text=element_text(size = 10))

snv_cdiff_plot

#There is one sample that has a very intensive peak at 1100, so we will delete that sample for further analysis. We consider it an outlier (removed after baseline correction)

#B) BASELINE CORRECTION --> Applied to SNV corrected data
ir_cdiff_bsln <- data.frame()
for (i in 1:nrow(ir_cdiff_snv)){
  numeric_vector <- ir_cdiff_snv[i,]
  row_df <- data.frame(matrix(baseline(numeric_vector), nrow = 1, ncol = length(numeric_vector)))
  ir_cdiff_bsln <- rbind(ir_cdiff_bsln,row_df)
}

#Now remove row 19, which we consider an outlier
ir_cdiff_bsln <- ir_cdiff_bsln[-19, ] # Outlier removal
colnames(ir_cdiff_bsln) <- colnames(ir_cdiff) #Use original column names
rownames(ir_cdiff_bsln) <- rownames(ir_cdiff[-19, ]) #Use original row names (excluding row 19)

#Remove it from SCFA data
scfa_cdiff <- scfa_cdiff[-19, ] # Remove outlier

#Data sorting (BASELINE CORRECTED DATA)
bsln_data_cdiff <- ir_cdiff_bsln %>% 
  rownames_to_column(var = "sample_id") %>%  #Row names to column
  pivot_longer(cols = -sample_id,            #Select all columns except "sample_id"
               names_to = "wavelength",     #Name for the wavelength column
               values_to = "absorbance")    #Name for the absorbance values

#Remove the 'X' before the wavelength value 
bsln_data_cdiff$wavelength <- gsub("X", "", bsln_data_cdiff$wavelength)  
bsln_data_cdiff$wavelength <- as.numeric(bsln_data_cdiff$wavelength)

#SNV + baseline correction plot (per sample) --> Check data visualization
blsn_cdiff_plot <- ggplot(bsln_data_cdiff, aes(x = wavelength, y = absorbance, color = sample_id)) + 
  geom_line(linewidth = 0.1) + 
  geom_hline(yintercept=0.0, linetype="dashed",color = "red", linewidth = 0.2) +
  labs(x = expression("Wavenumber (cm"^-1*")"), y = "Absorbance (a.u.)", title = "SNV + baseline corrected spectra - Healthy donors") + 
  scale_x_continuous(breaks = seq(min(bsln_data_cdiff$wavelength), max(bsln_data_cdiff$wavelength), by = 200), labels = seq(min(bsln_data_cdiff$wavelength), max(bsln_data_cdiff$wavelength), by = 200), expand = c(0, 0), trans = "reverse") +
  theme_minimal() +
  theme(text = element_text(size = 10),
        axis.title.y = element_text(vjust = +3, size = 12), 
        axis.title.x = element_text(vjust = +1, size = 12), 
        axis.ticks.x=element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none", 
        legend.text=element_text(size = 10))

blsn_cdiff_plot

#C) SAVITZKY-GOLAY --> 2nd derivate
ir_cdiff_sgk <- data.frame()
for (i in 1:nrow(ir_cdiff_bsln)){
  numeric_vector <- ir_cdiff_bsln[i,]
  row_df <- data.frame(savitzkyGolay(X = numeric_vector, m = 2, p = 3, w = 9))
  ir_cdiff_sgk <- rbind(ir_cdiff_sgk,row_df)
}

#Remove the X before each wavelenght
colnames(ir_cdiff_sgk) <- gsub("^X", "", colnames(ir_cdiff_sgk))

#Data sorting (SAVITZKY-GOLAY CORRECTED DATA)
skg_data_cdiff <- ir_cdiff_sgk %>% 
  rownames_to_column(var = "sample_id") %>%  #Row names to column
  pivot_longer(cols = -sample_id,            #Select all columns except "sample_id"
               names_to = "wavelength",     #Name for the wavelength column
               values_to = "absorbance")    #Name for the absorbance values

#Remove the 'X' before the wavelength value 
skg_data_cdiff$wavelength <- gsub("X", "", skg_data_cdiff$wavelength)  
skg_data_cdiff$wavelength <- as.numeric(skg_data_cdiff$wavelength)
```

## ABSORBANCE PLOTS

```{r}
#HEALTHY DONORS
#Add tag to indicate the patient type 
bsln_data_hd <- bsln_data_hd %>% mutate(patient_type = "healthy_donor")

#C. difficile PATIENTS
#Add tag to indicate the patient type 
bsln_data_cdiff <- bsln_data_cdiff %>% mutate(patient_type = "cdiff_patient")

#MERGE THE TWO DATAFRAMES 
corrected_ir_data <- rbind(bsln_data_hd, bsln_data_cdiff)

#MERGE THE TWO DATAFRAMES 
corrected_ir_avg <- corrected_ir_data %>%
  group_by(patient_type, wavelength) %>%
  summarize(absorbance_avg = mean(absorbance, na.rm = TRUE), .groups = "drop")
```

### Average healthy vs. CDI (Fig. 1B)

```{r}
#TOTAL PLOT (SNV + baseline correction plot (average))
plot_bonds <- ggplot(corrected_ir_avg, aes(x = wavelength, y = absorbance_avg, colour = patient_type)) + 
  geom_line(size = 0.8) + 
  scale_colour_manual(values = c("healthy_donor" = "skyblue4", "cdiff_patient" = "indianred4"), name = NULL, labels = c("Dysbiotic patients (n = 39)", "Healthy donors (n = 115)")) + 
  labs(x = expression("Wavenumber (cm"^-1*")"), y = "Absorbance (a.u.)") +
  annotate("rect", xmin=3600, xmax=3100, ymin=0, ymax=Inf, alpha=0.15, fill="red2") + #O-H region
  annotate("text", x = 3400, y = 3.5, label = paste0("O-H"), color = "red4", size = 4, hjust = 0) + #O-H letters
  annotate("rect", xmin=3000, xmax=2800, ymin=0, ymax=Inf, alpha=0.15, fill="steelblue4") + #CH region
annotate("text", x = 2900, y = 3.35, label = expression(atop(CH[2], CH[3])), color = "royalblue4", size = 4, hjust = 0.5) + #CH2/CH3 letters
  annotate("rect", xmin=1700, xmax=1400, ymin=0, ymax=Inf, alpha=0.15, fill="green4") + #C=O region 
  annotate("text", x = 1550, y = 3.5, label = paste0("C=O"), color = "darkgreen", size = 4, hjust = 0.5) + #C=O letters
  annotate("rect", xmin=1380, xmax=1210, ymin=0, ymax=Inf, alpha=0.15, fill="orange2") + #C-O region 
    annotate("text", x = 1300, y = 3.5, label = paste0("C-O"), color = "gold4", size = 4, hjust = 0.5) + #C=O letters
  scale_x_continuous(breaks = seq(min(corrected_ir_data$wavelength), max(corrected_ir_data$wavelength), by = 200), labels = seq(min(corrected_ir_data$wavelength), max(corrected_ir_data$wavelength), by = 200), expand = c(0, 0), trans = "reverse") +
  scale_y_continuous(limits = c(0, 3.75), breaks = c(0, 1, 2, 3)) +
    theme_minimal() +
  theme(text = element_text(size = 9.5),
        axis.title.y = element_text(vjust = +3, size = 11.5), 
        axis.title.x = element_text(vjust = +1, size = 11.5), 
        axis.ticks.x=element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom", 
        legend.text=element_text(size = 9.5))
  
plot_bonds

#ggsave("Figures_paper/Fig1B.Spectra_regions.jpeg", width = 6, height = 4, units = "in", dpi = 600)
```

### Savitzky-Golay (Supp. Fig. S1)

```{r}
#HEALTHY DONORS
#Add tag to indicate the patient type 
skg_data_hd <- skg_data_hd %>% mutate(patient_type = "healthy_donor")

#C. difficile PATIENTS
#Add tag to indicate the patient type 
skg_data_cdiff <- skg_data_cdiff %>% mutate(patient_type = "cdiff_patient")

#MERGE THE TWO DATAFRAMES 
skg_total_avg <- rbind(skg_data_hd, skg_data_cdiff)

#MERGE THE TWO DATAFRAMES 
skg_total_avg <- skg_total_avg %>%
  group_by(patient_type, wavelength) %>%
  summarize(absorbance_avg = mean(absorbance, na.rm = TRUE), .groups = "drop")

filtered_skg_average <- skg_total_avg %>% dplyr::filter(wavelength >= 500 & wavelength <= 3900)

#TOTAL PLOT (SNV + baseline correction plot (per sample))
skg_avg_plot <- ggplot(filtered_skg_average, aes(x = wavelength, y = absorbance_avg, color = patient_type)) + 
  geom_line(size = 0.4) + 
  scale_colour_manual(values = c("healthy_donor" = "skyblue4", "cdiff_patient" = "indianred4"), name = NULL, labels = c("Dysbiotic patients (n = 39)", "Healthy donors (n = 115)")) +   
  labs(x = expression("Wavenumber (cm"^-1*")"), y = "Second derivative values", title = "Savitzky-Golay - 2nd derivate") +
  annotate("rect", xmin=3600, xmax=3100, ymin=-0.010, ymax=Inf, alpha=0.15, fill="red2") + #O-H region
  annotate("text", x = 3400, y = 0.0094, label = paste0("O-H"), color = "red4", size = 4, hjust = 0) + #O-H letters
  annotate("rect", xmin=3000, xmax=2800, ymin=-0.010, ymax=Inf, alpha=0.15, fill="steelblue4") + #CH region
annotate("text", x = 2900, y = 0.008, label = expression(atop(CH[2], CH[3])), color = "royalblue4", size = 4, hjust = 0.5) + #CH2/CH3 letters
  annotate("rect", xmin=1700, xmax=1400, ymin=-0.010, ymax=Inf, alpha=0.15, fill="green4") + #C=O region 
  annotate("text", x = 1550, y = 0.0094, label = paste0("C=O"), color = "darkgreen", size = 4, hjust = 0.5) + #C=O letters
  annotate("rect", xmin=1380, xmax=1210, ymin=-0.010, ymax=Inf, alpha=0.15, fill="orange2") + #C-O region 
    annotate("text", x = 1300, y = 0.0094, label = paste0("C-O"), color = "gold4", size = 4, hjust = 0.5) + #C=O letters
  scale_x_continuous(breaks = seq(500, 3900, by = 200), expand = c(0, 0), trans = "reverse") +
  scale_y_continuous(limits = c(-0.010, 0.01), breaks = c(-0.010, -0.005, 0.000, 0.005)) +
  theme_minimal() +
  theme(text = element_text(size = 9.5),
      axis.title.y = element_text(vjust = +3, size = 11.5), 
      axis.title.x = element_text(vjust = +1, size = 11.5), 
      axis.ticks.x=element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "bottom", 
      legend.text=element_text(size = 9.5))
  
skg_avg_plot

#ggsave("Figures_paper/SupFigS1.Savitsky-Golay.jpeg", width = 6, height = 4, units = "in", dpi = 600)
```

## PCA - PRINCIPAL COMPONENT ANALYSIS

For the PCA analysis we need (FactoMineR package):

X = a data frame with *n* rows (individuals) and *p* columns (numeric variables)

ncp = number of dimensions kept in the results (by default 5)

quali.sup = column of qualitative data (for example patient type, in this case is column 1)

(<http://www.sthda.com/english/wiki/wiki.php?id_contents=7851#supplementary-qualitative-variables>)

### PCA - Selected regions (Fig. 1A)

```{r}
#Merge data from Savitzky-Golay (2nd derivate)
#For the PCA analysis we need a single dataframe with n rows (individuals) and p columns (numeric variables)
ir_skg_total <- rbind(ir_healthy_skg, ir_cdiff_sgk) %>%
  mutate(patient_type = rep(c("Healthy donors", "Dysbiotic patients"), times = c(115, 39))) %>%
  dplyr::select(patient_type, everything()) # Places `patient_type` at the begining

#write.csv(ir_skg_total, "ir_skg_data.csv")

#Selected columns according to groups and bonds in SCFA (O-H, C-H, C=O, C-O)
which(colnames(ir_skg_total) == "1210") #To check the column of the variable
ir_filtered <- ir_skg_total %>% dplyr::select(1, 394:894, 994:1194, 2294:2594, 2614:2784)

#PCA
#The variable 'patient_type' (column = 1) is removed before the PCA analysis
pca_skg_regions <- PCA(ir_filtered, ncp = 5, scale.unit = TRUE, quali.sup = 1)

#OUTPUTS PCA
summary(pca_skg_regions)  #Summary of PCA

#Eigenvalues --> a matrix containing all the eigenvalues, the percentage of variance and the cumulative percentage of variance
pca_skg_regions$eig       

#Coordinates (loadings) --> a list of matrices containing all the results for the active variables (coordinates, correlation between variables and axes, square cosine, contributions)
pca_skg_regions$var$coord 

#Individual scores --> a list of matrices containing all the results for the active individuals (coordinates, square cosine, contributions)
comp_ind <- pca_skg_regions$ind$coord 

#Check how many eigenvalues follow the Kaiser's criterium
sum(pca_skg_regions$eig[, 1] > 1)

#Plotting the PCA results with fviz_pca_ind (part of the factoextra package), uses ggplot2 internally 
pca_sgk_regions <- fviz_pca_ind(pca_skg_regions,  
             label="none", 
             habillage = 1, #Categorical data that separates results, here 'patient_type'
             addEllipses = TRUE, 
             ellipse.level = 0.95,
             legend.title = "Patient Type", 
             title = "SELECTED REGIONS - SCFAs",
             axes = c(1, 2), # Display PC1 and PC2
             xlab = paste0("PC1 (", round(pca_skg_regions$eig[1, 2], 2), "%)"), #PC1 
             ylab = paste0("PC2 (", round(pca_skg_regions$eig[2, 2], 2), "%)")) + #PC2
  scale_colour_manual(values = c("Healthy donors" = "skyblue4", "Dysbiotic patients" = "indianred4")) + 
  theme(text = element_text(size = 10),  
        axis.title.y = element_text(vjust = +2, size = 11),
        axis.title.x = element_text(vjust = +2, size = 11),
        legend.position = "bottom",  
        legend.box = "vertical",  # Stack legends vertically  
        legend.text = element_text(size = rel(0.8)))

pca_sgk_regions

#ggsave("Figures_paper/Fig1A.PCA_regions.jpeg", width = 6, height = 4, units = "in", dpi = 600)
```

#### **Scree plot Savitsky-Golay**

The proportion of variances retained by the principal components can be extracted.

-   **Eigenvalues**: Correspond to the amount of the variation explained by each principal component (PC). Eigenvalues are large for the first PC and small for the subsequent PCs.

-   A PC with an eigenvalue \> 1 indicates that the PC accounts for more variance than accounted by one of the original variables in standardized data. This is commonly used as a cutoff point to determine the number of PCs to retain.

```{r}
#Extract the eigenvalues
eigenvalues <- pca_skg_regions$eig
head(eigenvalues[, 1:2])

#Convert to a dataframe
eigenvalues_df <- data.frame(principal_component = 1:nrow(eigenvalues),
  variance_explained = eigenvalues[, 2])

#Scree plot 
scree_plot <- ggplot(eigenvalues_df, aes(x = principal_component, y = variance_explained)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.8) +
  geom_line(aes(group = 1), color = "red", size = 0.2) +
  geom_point(color = "red", size = 0.2) +
  geom_hline(yintercept=1, linetype="dashed",color = "red", linewidth = 0.2) +
  labs(title = "Scree Plot",
    x = "Principal Components",
    y = "Percentage of Variance Explained") +
  theme_minimal()

scree_plot

#Scree plot of the fifty first PC
#Filter eigenvalue > 1 indicates that the PC accounts for more variance than accounted by one of the original variables in standardized data. This is commonly used as a cutoff point to determine the number of PCs to retain.eigenvalues_df_50 <- eigenvalues_df %>%

signif_eigenvalues <- eigenvalues_df[eigenvalues_df$variance_explained >= 1, ]

scree_plot_signif <- ggplot(signif_eigenvalues, aes(x = principal_component, y = variance_explained)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.8) +
  geom_line(aes(group = 1), color = "red", size = 0.4) +
  geom_point(color = "red", size = 0.4) +
  geom_hline(yintercept=1, linetype="dashed",color = "red", linewidth = 0.2) +
  labs(title = "Scree Plot",
    x = "Principal Components",
    y = "Percentage of Variance Explained") +
  theme_minimal()
```

#### **Contribution of variables (Supp Fig. S3)**

The function fviz_contrib() creates a barplot of row/column contributions. A reference dashed line is also shown on the barplot. This reference line corresponds to the expected value if the contribution where uniform.

For a given dimension, any row/column with a contribution above the reference line could be considered as important in contributing to the dimension.

```{r}
#VARIABLE CONTRIBUTIONS
#Get the contribution of each variable to PC1 and PC2
var_contrib_reg <-as.data.frame(pca_skg_regions$var$contrib)  #Contributions of variables
head(var_contrib_reg)  #Show the first variables

#Visualizing Variable Contributions (1174 are all the variables used corresponding to the selected regions of SCFAs)

#PC1 - All variables 
all_variables_PC1 <- fviz_contrib(pca_skg_regions, choice = "var", axes = 1, top = 1174) +  # Var. contrib. to PC1
  theme_minimal() + 
  labs(title = "Contributions of Variables to PC1") + 
  theme(axis.title.x = element_blank(), axis.text.x = element_blank()) 

all_variables_PC1

#ggsave("Figures_paper/SupFigS3A.PC1_variables.jpeg", width = 6, height = 4, units = "in", dpi = 600)

#PC1 - First 200 Variable Contributions 
top_variables_PC1 <- fviz_contrib(pca_skg_regions, choice = "var", axes = 1, top = 100) + 
  theme_minimal() + 
  labs(title = "Contributions of top 100 Variables to PC1") + 
  theme(axis.title.x = element_blank(), axis.text.x = element_text(angle=45)) 

top_variables_PC1

#PC2 - All variables 
all_variables_PC2 <- fviz_contrib(pca_skg_regions, choice = "var", axes = 2, top = 1174) +  # Var. contrib. to PC2
  theme_minimal() + 
  labs(title = "Contributions of Variables to PC2") + 
  theme(axis.title.x = element_blank(), axis.text.x = element_blank()) 

all_variables_PC2

#ggsave("Figures_paper/SupFigS3B.PC2_variables.jpeg", width = 6, height = 4, units = "in", dpi = 600)

#PC2 - First 200 Variable Contributions 
top_variables_PC2 <- fviz_contrib(pca_skg_regions, choice = "var", axes = 2, top = 100) + 
  theme_minimal() + 
  labs(title = "Contributions of top 100 Variables to PC2") + 
  theme(axis.title.x = element_blank(), axis.text.x = element_text(angle=45)) 

top_variables_PC2
```

#### Mahalanobis distance - Decision rule for dysbiosis screening

```{r}
#MAHALANOBIS DISTANCE - DECISION RULE FOR DYSBIOSIS SCREENING
#1. Select PCs to use
#Kaiser criterion (eigenvalue > 1), capped at ncp stored in the PCA object.
#PCA was run with ncp = 5, so ind$coord has at most 5 columns.
n_pcs_kaiser <- sum(pca_skg_regions$eig[, 1] > 1)
n_pcs_stored <- ncol(pca_skg_regions$ind$coord)
n_pcs <- min(n_pcs_kaiser, n_pcs_stored)

#We choose to use only the first 2 PCs for better visualization and interpretability, even if more PCs have eigenvalue > 1.
n_pcs <- 2 
cat(sprintf("PCs with eigenvalue > 1: %d | PCs stored (ncp): %d | PCs used: %d\n",
            n_pcs_kaiser, n_pcs_stored, n_pcs))

#Extract PC scores for all samples
pca_scores <- as.data.frame(pca_skg_regions$ind$coord[, 1:n_pcs, drop = FALSE])
patient_type <- ir_filtered$patient_type

#2. Compute Mahalanobis distance to the healthy centroid 
healthy_scores <- pca_scores[patient_type == "Healthy donors", , drop = FALSE]

#Centroid and covariance of the healthy group in PC space
healthy_centroid <- colMeans(healthy_scores)
healthy_cov      <- cov(healthy_scores)

#Mahalanobis distance for every sample to the healthy centroid
mahal_distances <- mahalanobis(pca_scores, center = healthy_centroid, cov = healthy_cov)

mahal_df <- data.frame(
  sample_id      = rownames(ir_filtered),
  mahal_distance = mahal_distances,
  patient_type   = patient_type,
  is_dysbiotic   = ifelse(patient_type == "Dysbiotic patients", 1, 0)
)
```

##### ROC analysis (Supp. Fig S8)

```{r}
#3. ROC analysis to find the optimal distance threshold 
# direction = "<": lower distance → healthier → negative class
roc_obj <- roc(mahal_df$is_dysbiotic, mahal_df$mahal_distance,
               direction = "<", quiet = TRUE)

#Youden's J maximises sensitivity + specificity.
#For FMT screening, sensitivity is prioritised (avoid passing a dysbiotic donor).
cutoff_coords <- coords(roc_obj, "best", best.method = "youden",
                        ret = c("threshold", "sensitivity", "specificity", "accuracy"))

optimal_threshold   <- cutoff_coords$threshold
optimal_sensitivity <- cutoff_coords$sensitivity
optimal_specificity <- cutoff_coords$specificity

#4. Apply decision rule and report performance
mahal_df$prediction <- ifelse(mahal_df$mahal_distance < optimal_threshold,
                              "Healthy", "Dysbiotic")

confusion_matrix <- table(
  True      = ifelse(mahal_df$is_dysbiotic == 1, "Dysbiotic", "Healthy"),
  Predicted = mahal_df$prediction
)

overall_accuracy <- mean(mahal_df$prediction ==
                         ifelse(mahal_df$is_dysbiotic == 1, "Dysbiotic", "Healthy"))

cat("\n=== DYSBIOSIS SCREENING DECISION RULE SUMMARY ===\n")
cat(sprintf("PCs used: %d  |  Optimal Mahalanobis threshold: %.3f\n",
            n_pcs, optimal_threshold))
cat(sprintf("Sensitivity : %.1f%%\n", optimal_sensitivity * 100))
cat(sprintf("Specificity : %.1f%%\n", optimal_specificity * 100))
cat(sprintf("Accuracy    : %.1f%%\n", overall_accuracy   * 100))
cat(sprintf("AUC-ROC     : %.3f\n",   auc(roc_obj)))
cat("\nConfusion matrix:\n")
print(confusion_matrix)
cat("\nDecision rule:\n")
cat(sprintf("  Mahalanobis distance < %.3f  →  HEALTHY  (acceptable for FMT)\n", optimal_threshold))
cat(sprintf("  Mahalanobis distance >= %.3f  →  DYSBIOTIC (exclude from FMT)\n", optimal_threshold))

#5. ROC curve plot
roc_plot <- ggroc(roc_obj, colour = "steelblue4", linewidth = 1) +
  geom_abline(intercept = 1, slope = 1,
              linetype = "dashed", colour = "grey40", linewidth = 0.4) +
  annotate("text", x = 0.24, y = 0.24,
           label = paste0("AUC = ",  round(auc(roc_obj), 3),
                          "\nThreshold = ", round(optimal_threshold, 3),
                          "\nSens = ",      round(optimal_sensitivity, 3),
                          "\nSpec = ",      round(optimal_specificity, 3)),
           size = 4, color = "black", hjust = 0) +
  labs(title = "ROC Curve - Mahalanobis Distance",
       x = "1 - Specificity", y = "Sensitivity") +
  theme_minimal(base_size = 13)

roc_plot

#ggsave("Figures_paper/SupFigS8.ROC_Mahalanobis.jpeg", width = 6, height = 4, units = "in", dpi = 600)
```

##### PC scatter plot predicted class (Supp. Fig S9)

```{r}
#6. PC scatter plot coloured by predicted class
plot_df <- data.frame(
  PC1         = pca_scores[, 1],
  PC2         = pca_scores[, 2],
  true_status = patient_type,
  prediction  = mahal_df$prediction
)

plot_df$misclassified <- plot_df$prediction != 
  ifelse(plot_df$true_status == "Healthy donors", "Healthy", "Dysbiotic")

#Plot
decision_plot <- ggplot(plot_df,
                        aes(x = PC1, y = PC2,
                            colour = prediction, shape = true_status)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_point(data = plot_df %>% filter(misclassified),
             aes(fill = "Misclassified"),
             shape = 21, size = 5, colour = "orangered",
             stroke = 0.8) +           # no alpha here
  scale_colour_manual(values = c("Healthy"   = "skyblue4",
                                 "Dysbiotic" = "indianred4")) +
  scale_shape_manual(values = c("Healthy donors"     = 16,
                                "Dysbiotic patients" = 17)) +
  scale_fill_manual(name = "", values = c("Misclassified" = "transparent")) +
  guides(colour = guide_legend(order = 1),
         shape  = guide_legend(order = 2),
         fill   = guide_legend(order = 3,
                               override.aes = list(colour  = "orangered",
                                                   shape   = 1,
                                                   size    = 5,
                                                   stroke  = 0.8,
                                                   fill    = NA))) +
  labs(x      = paste0("PC1 (", round(pca_skg_regions$eig[1, 2], 2), "%)"),
       y      = paste0("PC2 (", round(pca_skg_regions$eig[2, 2], 2), "%)"),
       title  = "Decision Rule: Mahalanobis Distance to Healthy Centroid",
       colour = "Prediction", shape = "True Status") +
  theme(text = element_text(size = 7),  
        axis.title.y = element_text(vjust = +2, size = 7),
        axis.title.x = element_text(vjust = +2, size = 7),
        legend.position = "bottom",  
        legend.text = element_text(size = rel(0.8)))

decision_plot

#ggsave("Figures_paper/SupFigS9.Decision_plot.jpeg", width = 6, height = 4, units = "in", dpi = 600)
```

#### Spearman's correlation: MD vs SCFAs

```{r}
#SCFA + FT-IR MAHALANOBIS DISTANCE SCORES
scfa_total <- scfa_total %>% remove_rownames %>% column_to_rownames(var='sample_id')

mahal_scfa <- merge(mahal_df, scfa_total, by = "row.names", all = FALSE) #Join by sample id, in this case is the row name
rownames(mahal_scfa) <- mahal_scfa$Row.names #Column "row.names" back to indeed row names
mahal_scfa <- mahal_scfa[, -1]  #Drop the Row.names column

#Correlation for between the Mahalanobis distance and each SCFAs
cor_acetate <- cor.test(mahal_scfa$mahal_distance, mahal_scfa$acetate, method = "spearman")
cor_propionate <- cor.test(mahal_scfa$mahal_distance, mahal_scfa$propionate,     method = "spearman")
cor_butyrate <- cor.test(mahal_scfa$mahal_distance, mahal_scfa$butyrate,     method = "spearman")

#Print correlation results
cat("\n--- Spearman correlations ---\n")
cat("Acetate vs MD:     rho =", cor_acetate$estimate,    "p =", cor_acetate$p.value, "\n")
cat("Propionate vs MD:  rho =", cor_propionate$estimate, "p =", cor_propionate$p.value, "\n")
cat("Butyrate vs MD:    rho =", cor_butyrate$estimate,   "p =", cor_butyrate$p.value, "\n")
```

#### LDA nested LOOCV + Bootstrap confidence intervals

[KEY DESIGN DECISIONS:]{.underline}

1.  **Nested LOOCV:** PCA is recomputed from scratch on N-1 training samples each fold. The left-out sample is projected onto the TRAINING PCA space. ---\> This eliminates data leakage compared to the approach where we just remove the data point sample from the PCA calculations.

2.  **Bootstrap CI (B=1000):** each re-sample runs the full PCA+LDA pipeline on the bootstrap sample and evaluates on out-of-bag (OOB) samples. ---\> This directly addresses sample-composition robustness.

3.  Equal priors (0.5/0.5) throughout to correct for 25/75 class imbalance.

```{r}
#LDA - NESTED LOOCV + BOOTSTRAP CI

#Input: raw spectral matrix (no patient_type column) + labels
#We use ir_filtered directly so that PCA is refit inside each fold/bootstrap
spectral_data <- as.matrix(ir_filtered[, -1])   # drop patient_type column
patient_type  <- ir_filtered$patient_type
group_factor  <- factor(patient_type)
n_samples     <- nrow(spectral_data)

cat("Class distribution:
")
print(table(group_factor))

#Helper: given train/test spectral matrices + labels, fit PCA+LDA and return
#P(Dysbiotic) for each test sample. npc = number of PCs to retain.
run_pca_lda <- function(train_X, test_X, train_y, npc) {
  #PCA on training data only
  pca_fit    <- prcomp(train_X, center = TRUE, scale. = TRUE)
  train_pcs  <- as.data.frame(pca_fit$x[, 1:npc, drop = FALSE])
  test_pcs   <- as.data.frame(predict(pca_fit, newdata = test_X)[, 1:npc, drop = FALSE])

  train_df   <- data.frame(train_pcs, group = train_y)

  lda_fit    <- lda(group ~ ., data = train_df, prior = c(0.5, 0.5))
  lda_pred   <- predict(lda_fit, newdata = test_pcs)

  list(
    prob  = lda_pred$posterior[, "Dysbiotic patients"],
    class = as.character(lda_pred$class)
  )
}

#Number of PCs: use Kaiser criterion on the FULL PCA (for reference),
#capped at ncp=5 stored in the existing pca_skg_regions object.
#Inside each fold prcomp retains all components; we slice to npc.
npc <- min(sum(pca_skg_regions$eig[, 1] > 1), 5)
cat(sprintf("
PCs used in nested CV: %d
", npc))

#1. NESTED LOOCV
loocv_probs  <- numeric(n_samples)
loocv_labels <- character(n_samples)

for (i in seq_len(n_samples)) {
  train_X <- spectral_data[-i, , drop = FALSE]
  test_X  <- spectral_data[ i, , drop = FALSE]
  train_y <- group_factor[-i]

  res <- run_pca_lda(train_X, test_X, train_y, npc)
  loocv_probs[i]  <- res$prob
  loocv_labels[i] <- res$class
}

loocv_labels <- factor(loocv_labels, levels = levels(group_factor))

#Performance
cm_loocv <- confusionMatrix(loocv_labels, group_factor,
                             positive = "Dysbiotic patients")
roc_loocv <- roc(ifelse(group_factor == "Dysbiotic patients", 1, 0),
                 loocv_probs, direction = "<", quiet = TRUE)

#MCC
tp <- cm_loocv$table["Dysbiotic patients", "Dysbiotic patients"]
tn <- cm_loocv$table["Healthy donors",     "Healthy donors"]
fp <- cm_loocv$table["Dysbiotic patients", "Healthy donors"]
fn <- cm_loocv$table["Healthy donors",     "Dysbiotic patients"]
mcc_num <- (tp * tn) - (fp * fn)
mcc_den <- sqrt((tp + fp) * (tp + fn) * (tn + fp) * (tn + fn))
mcc_obs <- ifelse(mcc_den == 0, 0, mcc_num / mcc_den)

cat("
=== NESTED LOOCV PERFORMANCE (no data leakage) ===
")
cat(sprintf("AUROC             : %.3f
",  as.numeric(auc(roc_loocv))))
cat(sprintf("Sensitivity       : %.1f%%
", cm_loocv$byClass["Sensitivity"]       * 100))
cat(sprintf("Specificity       : %.1f%%
", cm_loocv$byClass["Specificity"]       * 100))
cat(sprintf("Balanced Accuracy : %.1f%%
", cm_loocv$byClass["Balanced Accuracy"] * 100))
cat(sprintf("MCC               : %.3f
",  mcc_obs))
cat("
Confusion matrix:
")
print(cm_loocv$table)
```

##### ROC plot - Nested LOOCV (Supp. Fig S10)

```{r}
#ROC plot — nested LOOCV
roc_nested_plot <- ggroc(roc_loocv, colour = "steelblue4", linewidth = 1) +
  geom_abline(intercept = 1, slope = 1,
              linetype = "dashed", colour = "grey40", linewidth = 0.4) +
  annotate("text", x = 0.24, y = 0.24,
           label = paste0("AUC = ",            round(auc(roc_loocv), 3),
                          "
Sens = ",         round(cm_loocv$byClass["Sensitivity"],       3),
                          "
Spec = ",         round(cm_loocv$byClass["Specificity"],       3),
                          "
Bal. Acc = ",     round(cm_loocv$byClass["Balanced Accuracy"], 3),
                          "
MCC = ",          round(mcc_obs, 3)),
           size = 4, colour = "black", hjust = 0) +
  labs(title = "ROC Curve - LDA (Nested LOOCV, no leakage)",
       x = "1 - Specificity", y = "Sensitivity") +
  theme_minimal(base_size = 13)

roc_nested_plot

#ggsave("Figures_paper/SupFigS10.ROC_LOOCV-LDA.jpeg", width = 6, height = 4, units = "in", dpi = 600)
```

##### Bootstrap (Supp. Fig. S11 and S12)

```{r}
#2. BOOTSTRAP CONFIDENCE INTERVALS (B = 1000)
# Each bootstrap iteration:
#   a) Sample N indices WITH replacement -> bootstrap sample
#   b) Out-of-bag (OOB) = samples NOT selected (typically ~37% of N)
#   c) Fit PCA + LDA on bootstrap sample
#   d) Predict on OOB samples -> compute metrics
# This tests robustness to different sample compositions.

set.seed(123)
B <- 1000

boot_auc   <- numeric(B)
boot_sens  <- numeric(B)
boot_spec  <- numeric(B)
boot_bacc  <- numeric(B)
boot_mcc   <- numeric(B)

valid_boots <- 0  # count iterations with usable OOB sets

for (b in seq_len(B)) {
  boot_idx <- sample(seq_len(n_samples), replace = TRUE)
  oob_idx  <- setdiff(seq_len(n_samples), boot_idx)

  # Skip if OOB has fewer than 2 samples or missing a class
  if (length(oob_idx) < 2) next
  oob_classes <- group_factor[oob_idx]
  if (length(unique(oob_classes)) < 2) next

  train_X <- spectral_data[boot_idx, , drop = FALSE]
  test_X  <- spectral_data[oob_idx,  , drop = FALSE]
  train_y <- group_factor[boot_idx]

  res <- tryCatch(
    run_pca_lda(train_X, test_X, train_y, npc),
    error = function(e) NULL
  )
  if (is.null(res)) next

  pred_labels_b <- factor(res$class, levels = levels(group_factor))
  true_labels_b <- oob_classes

  cm_b <- tryCatch(
    confusionMatrix(pred_labels_b, true_labels_b, positive = "Dysbiotic patients"),
    error = function(e) NULL
  )
  if (is.null(cm_b)) next

  roc_b <- tryCatch(
    roc(ifelse(true_labels_b == "Dysbiotic patients", 1, 0),
        res$prob, direction = "<", quiet = TRUE),
    error = function(e) NULL
  )
  if (is.null(roc_b)) next

  tp_b <- cm_b$table["Dysbiotic patients", "Dysbiotic patients"]
  tn_b <- cm_b$table["Healthy donors",     "Healthy donors"]
  fp_b <- cm_b$table["Dysbiotic patients", "Healthy donors"]
  fn_b <- cm_b$table["Healthy donors",     "Dysbiotic patients"]
  den_b <- sqrt((tp_b + fp_b) * (tp_b + fn_b) * (tn_b + fp_b) * (tn_b + fn_b))

  valid_boots <- valid_boots + 1
  boot_auc[valid_boots]  <- as.numeric(auc(roc_b))
  boot_sens[valid_boots] <- cm_b$byClass["Sensitivity"]
  boot_spec[valid_boots] <- cm_b$byClass["Specificity"]
  boot_bacc[valid_boots] <- cm_b$byClass["Balanced Accuracy"]
  boot_mcc[valid_boots]  <- ifelse(den_b == 0, 0,
                                   ((tp_b * tn_b) - (fp_b * fn_b)) / den_b)
}

# Trim to valid iterations
boot_auc   <- boot_auc[1:valid_boots]
boot_sens  <- boot_sens[1:valid_boots]
boot_spec  <- boot_spec[1:valid_boots]
boot_bacc  <- boot_bacc[1:valid_boots]
boot_mcc   <- boot_mcc[1:valid_boots]

ci95 <- function(x) quantile(x, c(0.025, 0.975), na.rm = TRUE)

cat(sprintf("
=== BOOTSTRAP CONFIDENCE INTERVALS (B = %d valid resamples) ===
",
            valid_boots))
cat(sprintf("AUROC             : %.3f  [%.3f - %.3f]
",
            median(boot_auc),  ci95(boot_auc)[1],  ci95(boot_auc)[2]))
cat(sprintf("Sensitivity       : %.3f  [%.3f - %.3f]
",
            median(boot_sens), ci95(boot_sens)[1], ci95(boot_sens)[2]))
cat(sprintf("Specificity       : %.3f  [%.3f - %.3f]
",
            median(boot_spec), ci95(boot_spec)[1], ci95(boot_spec)[2]))
cat(sprintf("Balanced Accuracy : %.3f  [%.3f - %.3f]
",
            median(boot_bacc), ci95(boot_bacc)[1], ci95(boot_bacc)[2]))
cat(sprintf("MCC               : %.3f  [%.3f - %.3f]
",
            median(boot_mcc),  ci95(boot_mcc)[1],  ci95(boot_mcc)[2]))

# Bootstrap distribution plots
boot_results_df <- data.frame(
  AUC              = boot_auc,
  Sensitivity      = boot_sens,
  Specificity      = boot_spec,
  Balanced_Accuracy = boot_bacc,
  MCC              = boot_mcc
)

boot_long <- pivot_longer(boot_results_df, everything(),
                          names_to = "metric", values_to = "value")

# Observed values for vertical reference lines
obs_vals <- data.frame(
  metric = c("AUC", "Sensitivity", "Specificity", "Balanced_Accuracy", "MCC"),
  obs    = c(as.numeric(auc(roc_loocv)),
             cm_loocv$byClass["Sensitivity"],
             cm_loocv$byClass["Specificity"],
             cm_loocv$byClass["Balanced Accuracy"],
             mcc_obs)
)

# Boot_plot of MCC
boot_long_mcc <- boot_long %>% filter(metric == "MCC")
obs_vals_mcc <- obs_vals %>% filter(metric == "MCC")

# MCC range is from -1 to 1, so we adjust x-axis limits for better visualization
boot_plot_mcc <- ggplot(boot_long_mcc, aes(x = value)) +
  geom_histogram(bins = 35, fill = "grey70", colour = "white") +
  geom_vline(data = obs_vals_mcc, aes(xintercept = obs),
             colour = "indianred4", linewidth = 0.9, linetype = "solid") +
  scale_x_continuous(limits = c(-1, 1)) +  # Set x-axis limits from -1 to 1
  labs(title = "Bootstrap distribution of MCC (B=1000 OOB resamples)",
       subtitle = "Red line = observed nested LOOCV value",
       x = "MCC value", y = "Count") +
  theme_minimal(base_size = 12) +
  theme(strip.text = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))

boot_plot_mcc

#ggsave("Figures_paper/SupFigS11.Distribution_boot_MCC.jpeg", width = 6, height = 4, units = "in", dpi = 600)

# Updated boot_plot without MCC
boot_long_no_mcc <- boot_long %>% filter(metric != "MCC")
obs_vals_no_mcc <- obs_vals %>% filter(metric != "MCC")

boot_plot <- ggplot(boot_long_no_mcc, aes(x = value)) +
  geom_histogram(bins = 100, fill = "grey70", colour = "white") +
  geom_vline(data = obs_vals_no_mcc, aes(xintercept = obs),
             colour = "indianred4", linewidth = 0.9, linetype = "dashed") +
  geom_text(data = obs_vals_no_mcc, aes(x = 0.5, y = Inf, label = paste0("(", round(obs, 3), ")")),
            vjust = 1.5, hjust = 0.5, size = 3, colour = "indianred4", inherit.aes = FALSE) +
  scale_x_continuous(limits = c(0, 1.01)) +  # Set x-axis limits from 0 to 1
  facet_wrap(~ metric, ncol = 2) +
  labs(title = paste0("Bootstrap distributions of classification ",
                      "metrics (B=1000 OOB resamples)"),
       x = "Metric value", y = "Count") +
  theme_minimal(base_size = 10) +
  theme(strip.text = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))

boot_plot

#ggsave("Figures_paper/SupFigS12.Distribution_boot_metrics.jpeg", width = 6, height = 4, units = "in", dpi = 600)
```

### PCA - All regions (Supp. Fig. S2)

```{r}
#Merge data from Savitzky-Golay (2nd derivate)
#For the PCA analysis we need a single dataframe with n rows (individuals) and p columns (numeric variables)
ir_skg_total <- rbind(ir_healthy_skg, ir_cdiff_sgk) %>%
  mutate(patient_type = rep(c("Healthy donors", "Dysbiotic patients"), times = c(115, 39))) %>%
  dplyr::select(patient_type, everything()) # Places `patient_type` at the begining

#PCA
#The variable 'patient_type' (column = 1) is removed before the PCA analysis
pca_skg <- PCA(ir_skg_total, ncp = 5, scale = TRUE, quali.sup = 1)

#OUTPUTS PCA
summary(pca_skg)  #Summary of PCA

#Eigenvalues --> a matrix containing all the eigenvalues, the percentage of variance and the cumulative percentage of variance
pca_skg$eig       

#Coordinates (loadings) --> a list of matrices containing all the results for the active variables (coordinates, correlation between variables and axes, square cosine, contributions)
pca_skg$var$coord 

#Individual scores --> a list of matrices containing all the results for the active individuals (coordinates, square cosine, contributions)
comp_ind <- pca_skg$ind$coord 

#Plotting the PCA results with fviz_pca_ind (part of the factoextra package), uses ggplot2 internally 
pca_sgk <- fviz_pca_ind(pca_skg,  
             label="none", 
             habillage = 1, #Categorical data that separates results, here 'patient_type'
             addEllipses = TRUE, 
             ellipse.level = 0.95,
             legend.title = "Patient Type", 
             axes = c(1, 2), # Display PC1 and PC2
             xlab = paste0("PC1 (", round(pca_skg$eig[1, 2], 2), "%)"), #PC1 
             ylab = paste0("PC2 (", round(pca_skg$eig[2, 2], 2), "%)")) + #PC2
  scale_colour_manual(values = c("Healthy donors" = "gold3", "Dysbiotic patients" = "orchid4")) + 
  theme(text = element_text(size = 10),  
        axis.title.y = element_text(vjust = +2, size = 11),
        axis.title.x = element_text(vjust = +2, size = 11),
        legend.position = "bottom",  
        legend.box = "vertical",  # Stack legends vertically  
        legend.text = element_text(size = rel(0.8)))

pca_sgk

#ggsave("Figures_paper/SupFigS2.PCA_all_regions.jpeg", width = 6, height = 4, units = "in", dpi = 600)
```

### PCA - SCFAs levels

SCFA levels + IR data:

```{r}
#SCFA + IR DATA
scfa_total <- rbind(scfa_hd, scfa_cdiff)
scfa_total <- scfa_total %>% remove_rownames %>% column_to_rownames(var='sample_id')

#SCFA BOXPLOTS 
scfa_boxplots <- scfa_total %>% mutate(patient_type = rep(c("healthy_donors", "cdiff_patients"), times = c(115, 39)))

#Use the ir_filtered to ensure we are using the selected regions of SCFAs
ftir_total <- merge(ir_filtered, scfa_total, by = "row.names", all = FALSE) #Join by sample id, in this case is the row name
rownames(ftir_total) <- ftir_total$Row.names #Column "row.names" back to indeed row names
ftir_total <- ftir_total[, -1]  #Drop the Row.names column

ftir_total <- ftir_total %>% dplyr::select(-patient_type) %>% dplyr::select(acetate, propionate, butyrate, everything())  #Places the SCFAs at the begining
```

#### ACETATE (Supp. Fig. S4A and B)

PCA acetate (Supplementary Figure S4 A):

```{r}
#ACETATE
#Quartiles acetate
acetate_ftir <- ftir_total %>% #Select acetate
  mutate(acetate_quartiles = ntile(acetate, 10)) %>% #Divide 'acetate' into quartiles
  dplyr::select(acetate, acetate_quartiles, everything()) %>% 
  mutate(acetate_levels = case_when(acetate_quartiles == "10" ~ "High acetate",
                                    acetate_quartiles == "1" ~ "Low acetate",
                                    TRUE ~ "Other")) %>% #Qualitative tags
  dplyr::select(acetate_levels, everything()) %>%
  dplyr::select(-acetate_quartiles, -acetate, -propionate, -butyrate) #Drop SCFAs and patient type, we only want the quartiles

#NOTE!: The quartiles are made with all the data, not by groups (C.diff vs. Healthy)

#PCA - ACETATE LEVELS 
#The variable 'acetate_levels' (column = 1) is removed before the PCA analysis, is qualitative
pca_acetate <- PCA(acetate_ftir, ncp = 5, scale = TRUE, quali.sup = 1)

#OUTPUTS PCA
summary(pca_acetate)  #Summary of PCA

#Eigenvalues --> a matrix containing all the eigenvalues, the percentage of variance and the cumulative percentage of variance
pca_acetate$eig       

#Coordinates (loadings) --> a list of matrices containing all the results for the active variables (coordinates, correlation between variables and axes, square cosine, contributions)
pca_acetate$var$coord 

#Individual scores --> a list of matrices containing all the results for the active individuals (coordinates, square cosine, contributions)
comp_ind <- pca_acetate$ind$coord 

#Plotting the PCA results with fviz_pca_ind (part of the factoextra package), uses ggplot2 internally 
pca_sgk_acetate <- fviz_pca_ind(pca_acetate,  
             label="none", 
             habillage = 1, #Categorical data that separates results, here 'acetate levels'
             addEllipses = FALSE, 
             ellipse.level = 0.95,
             legend.title = "Acetate levels", 
             axes = c(1, 2), # Display PC1 and PC2
             xlab = paste0("PC1 (", round(pca_acetate$eig[1, 2], 2), "%)"), #PC1 
             ylab = paste0("PC2 (", round(pca_acetate$eig[2, 2], 2), "%)")) + #PC2
  scale_colour_manual(values = c("Other" = "gray80", "Low acetate" = "orange", "High acetate" = "forestgreen")) + 
  scale_shape_manual(values = c("Other" = 3, "Low acetate" = 17, "High acetate" = 15)) + 
  labs(title = NULL) +
  theme_minimal() +
  theme(text = element_text(size = 10),  
        axis.title.y = element_text(vjust = +2, size = 11),
        axis.title.x = element_text(vjust = +2, size = 11),
        legend.position = "bottom",  
        legend.text = element_text(size = rel(0.8)))

pca_sgk_acetate

#ggsave("Figures_paper/SupFigS4B.PCA_acetate.jpeg", width = 6, height = 4, units = "in", dpi = 600)
```

Boxplot acetate (Supplementary Figure S4 B):

```{r}
#ACETATE
#First, we need to add the column according to acetate levels (high or low). To do so:
#1. Move rownames to a "sample_id" variable in the two dataframes of interest
acetate_ftir <- acetate_ftir %>% rownames_to_column(var = "sample_id")

acetate_data <- scfa_boxplots %>% dplyr::select(acetate, patient_type) %>% rownames_to_column(var = "sample_id")

#2. Add the column of interest according to "sample_id"
acetate_data <- acetate_data %>%
  left_join(acetate_ftir %>% dplyr::select(sample_id, acetate_levels), by = "sample_id")
  
#3. Convert sample_id back to rownames in the dataframe used for the PCA to avoid errors
acetate_ftir <- acetate_ftir %>% column_to_rownames(var = "sample_id")

#VIOLIN PLOT ACETATE
plot_acetate <- ggplot(acetate_data, aes(x = patient_type, y = acetate, fill = patient_type)) +
  geom_violin(alpha = 0.2, width = 0.5) +  
  geom_boxplot(alpha = 0.3, width = 0.1, outliers = FALSE) +  
  scale_fill_manual(values = c("healthy_donors" = "skyblue4", "cdiff_patients" = "indianred4")) +
  geom_point(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.5, seed = 42), 
             aes(group = patient_type, colour = acetate_levels, shape = acetate_levels), 
             alpha = 0.75, size = 2.5) +  
  scale_colour_manual(values = c("High acetate" = "forestgreen", "Low acetate" = "orange", "Other" = "grey60"), name = NULL) +  
  scale_shape_manual(values = c("High acetate" = 15, "Low acetate" = 17, "Other" = 3)) +  
  scale_x_discrete(label = c("Dysbiotic patients", "Healthy donors")) +  
  scale_y_continuous(limits = c(0, 680), breaks = c(0, 200, 400, 600)) +
  labs(x = NULL, y = "Normalized acetate levels") +  
  guides(colour = guide_legend(order = 2, override.aes = list(shape = c(15, 17, 3))),
         shape = "none", fill = "none") +  
  theme_light() +  
  theme(text = element_text(size = 14),  
        axis.title.y = element_text(vjust = +3, size = 14),  
        legend.position = "bottom",  
        legend.box = "vertical",  # Stack legends vertically  
        legend.text = element_text(size = rel(0.8)))

plot_acetate

#ggsave("Figures_paper/SupFigS4A.Boxplot_acetate.jpeg", width = 6, height = 6, units = "in", dpi = 600)
```

#### PROPIONATE (Supp. Fig. S4C and D)

PCA propionate (Supplementary Figure S4 C):

```{r}
#PROPIONATE
#Quartiles propionate
propionate_ftir <- ftir_total %>% #Select propionate
  mutate(propionate_quartiles = ntile(propionate, 10)) %>% #Divide 'propionate' into quartiles
  dplyr::select(propionate, propionate_quartiles, everything()) %>%
  mutate(propionate_levels = case_when(propionate_quartiles == "10" ~ "High propionate",
                                     propionate_quartiles == "1" ~ "Low propionate", 
                                     TRUE ~ "Other")) %>% #Qualitative tags
  dplyr::select(propionate_levels, everything()) %>%
  dplyr::select(-propionate_quartiles, -acetate, -propionate, -butyrate) #Drop SCFAs and patient type, we only want the quartiles

#NOTE!: The quartiles are made with all the data, not by groups (C.diff vs. Healthy)

#PCA - PROPIONATE LEVELS 
#The variable 'propionate_levels' (column = 1) is removed before the PCA analysis, is qualitative
pca_propionate <- PCA(propionate_ftir, ncp = 5, scale = TRUE, quali.sup = 1)

#OUTPUTS PCA
summary(pca_propionate)  #Summary of PCA

#Eigenvalues --> a matrix containing all the eigenvalues, the percentage of variance and the cumulative percentage of variance
pca_propionate$eig       

#Coordinates (loadings) --> a list of matrices containing all the results for the active variables (coordinates, correlation between variables and axes, square cosine, contributions)
pca_propionate$var$coord 

#Individual scores --> a list of matrices containing all the results for the active individuals (coordinates, square cosine, contributions)
comp_ind <- pca_propionate$ind$coord 

#Plotting the PCA results with fviz_pca_ind (part of the factoextra package), uses ggplot2 internally 
pca_sgk_propionate <- fviz_pca_ind(pca_propionate,  
             label="none", 
             habillage = 1, #Categorical data that separates results, here 'acetate levels'
             addEllipses = FALSE, 
             ellipse.level = 0.95,
             legend.title = "Propionate levels", 
             axes = c(1, 2), # Display PC1 and PC2
             xlab = paste0("PC1 (", round(pca_propionate$eig[1, 2], 2), "%)"), #PC1 
             ylab = paste0("PC2 (", round(pca_propionate$eig[2, 2], 2), "%)")) + #PC2
  scale_colour_manual(values = c("Other" = "gray80", "Low propionate" = "orange", "High propionate" = "forestgreen")) + 
  scale_shape_manual(values = c("Other" = 3, "Low propionate" = 17, "High propionate" = 15)) +   labs(title = NULL) +
  theme_minimal() +
  theme(text = element_text(size = 10),  
      axis.title.y = element_text(vjust = +2, size = 11),
      axis.title.x = element_text(vjust = +2, size = 11),
      legend.position = "bottom",  
      legend.text = element_text(size = rel(0.8)))

pca_sgk_propionate

#ggsave("Figures_paper/SupFigS4D.PCA_propionate.jpeg", width = 6, height = 4, units = "in", dpi = 600)

#BEST PLOTS --> PC1 and PC2, and PC1 and PC3
```

Boxplot propionate (Supplementary Figure S4 D):

```{r}
#PROPIONATE
#First, we need to add the column according to propionate levels (high or low). To do so:
#1. Move rownames to a "sample_id" variable in the two dataframes of interest
propionate_ftir <- propionate_ftir %>% rownames_to_column(var = "sample_id")

propionate_data <- scfa_boxplots %>% dplyr::select(propionate, patient_type) %>% rownames_to_column(var = "sample_id")

#2. Add the column of interest according to "sample_id"
propionate_data <- propionate_data %>%
  left_join(propionate_ftir %>% dplyr::select(sample_id, propionate_levels), by = "sample_id")
  
#3. Convert sample_id back to rownames in the dataframe used for the PCA to avoid errors
propionate_ftir <- propionate_ftir %>% column_to_rownames(var = "sample_id")

#VIOLIN PLOT PROPIONATE
plot_propionate <- ggplot(propionate_data, aes(x = patient_type, y = propionate, fill = patient_type)) +
  geom_violin(alpha = 0.2, width = 0.5) +  
  geom_boxplot(alpha = 0.3, width = 0.1, outliers = FALSE) +  
  scale_fill_manual(values = c("healthy_donors" = "skyblue4", "cdiff_patients" = "indianred4")) +
  geom_point(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.5, seed = 42), 
             aes(group = patient_type, colour = propionate_levels, shape = propionate_levels), 
             alpha = 0.75, size = 2.5) +  
  scale_colour_manual(values = c("High propionate" = "forestgreen", "Low propionate" = "orange", "Other" = "grey60"), name = NULL) +  
  scale_shape_manual(values = c("High propionate" = 15, "Low propionate" = 17, "Other" = 3)) +  
  scale_x_discrete(label = c("Dysbiotic patients", "Healthy donors")) +  
  scale_y_continuous(limits = c(0, 300), breaks = c(0, 100, 200, 300)) +
  labs(x = NULL, y = "Normalized propionate levels") +  
  guides(colour = guide_legend(order = 2, override.aes = list(shape = c(15, 17, 3))),
         shape = "none", fill = "none") +  
  theme_light() +  
  theme(text = element_text(size = 14),  
        axis.title.y = element_text(vjust = +3, size = 14),  
        legend.position = "bottom",  
        legend.box = "vertical",  # Stack legends vertically  
        legend.text = element_text(size = rel(0.8)))

plot_propionate

#ggsave("Figures_paper/SupFigS4C.Boxplot_propionate.jpeg", width = 6, height = 6, units = "in", dpi = 600)
```

#### BUTYRATE (Fig. 2 A and B)

PCA butyrate (Fig 2A)

```{r}
#BUTYRATE
#Quartiles butyrate
butyrate_ftir <- ftir_total %>% #Select butyrate
 mutate(butyrate_quartiles = ntile(butyrate, 10)) %>%
  dplyr::select(butyrate, butyrate_quartiles, everything()) %>% 
  mutate(butyrate_levels = case_when(butyrate_quartiles == "10" ~ "High butyrate",
                                     butyrate_quartiles == "1" ~ "Low butyrate", 
                                     TRUE ~ "Other")) %>% #Qualitative tags
 dplyr::select(butyrate_levels, everything()) %>%
  dplyr::select(-butyrate_quartiles, -acetate, -propionate, -butyrate) #Drop SCFAs and patient type, we only want the quartiles

#NOTE!: The quartiles are made with all the data, not by groups (C.diff vs. Healthy)

#PCA - BUYTRATE LEVELS 
#The variable 'butyrate_levels' (column = 1) is removed before the PCA analysis, is qualitative
pca_butyrate <- PCA(butyrate_ftir, ncp = 5, scale = TRUE, quali.sup = 1)

#OUTPUTS PCA
summary(pca_butyrate)  #Summary of PCA

#Eigenvalues --> a matrix containing all the eigenvalues, the percentage of variance and the cumulative percentage of variance
pca_butyrate$eig       

#Coordinates (loadings) --> a list of matrices containing all the results for the active variables (coordinates, correlation between variables and axes, square cosine, contributions)
pca_butyrate$var$coord 

#Individual scores --> a list of matrices containing all the results for the active individuals (coordinates, square cosine, contributions)
comp_ind <- pca_butyrate$ind$coord 

#Plotting the PCA results with fviz_pca_ind (part of the factoextra package), uses ggplot2 internally 
pca_sgk_butyrate <- fviz_pca_ind(pca_butyrate,  
             label="none", 
             habillage = 1, #Categorical data that separates results, here 'acetate levels'
             addEllipses = FALSE, 
             ellipse.level = 0.95,
             legend.title = "Butyrate levels", 
             axes = c(1, 2), # Display PC1 and PC2
             xlab = paste0("PC1 (", round(pca_butyrate$eig[1, 2], 2), "%)"), #PC1 
             ylab = paste0("PC2 (", round(pca_butyrate$eig[2, 2], 2), "%)")) + #PC2
  scale_colour_manual(values = c("Other" = "gray80", "Low butyrate" = "orange", "High butyrate" = "forestgreen")) + 
  scale_shape_manual(values = c("Other" = 3, "Low butyrate" = 17, "High butyrate" = 15)) +   labs(title = NULL) +
  theme_minimal() +
  theme(text = element_text(size = 10),  
      axis.title.y = element_text(vjust = +2, size = 11),
      axis.title.x = element_text(vjust = +2, size = 11),
      legend.position = "bottom",  
      legend.text = element_text(size = rel(0.8)))

pca_sgk_butyrate

#ggsave("Figures_paper/Fig2B.PCA_butyrate.jpeg", width = 6, height = 4, units = "in", dpi = 600)

#BEST PLOTS --> PC1 and PC2, and PC1 and PC3
```

Boxplot butyrate (Fig 2B)

```{r}
#BUTYRATE
#First, we need to add the column according to butyrate levels (high or low). To do so:
#1. Move rownames to a "sample_id" variable in the two dataframes of interest
butyrate_ftir <- butyrate_ftir %>% rownames_to_column(var = "sample_id")

butyrate_data <- scfa_boxplots %>% dplyr::select(butyrate, patient_type) %>% rownames_to_column(var = "sample_id")

#2. Add the column of interest according to "sample_id"
butyrate_data <- butyrate_data %>%
  left_join(butyrate_ftir %>% dplyr::select(sample_id, butyrate_levels), by = "sample_id")
  
#3. Convert sample_id back to rownames in the dataframe used for the PCA to avoid errors
butyrate_ftir <- butyrate_ftir %>% column_to_rownames(var = "sample_id")

#VIOLIN PLOT BUTYRATE
plot_butyrate <- ggplot(butyrate_data, aes(x = patient_type, y = butyrate, fill = patient_type)) +
  geom_violin(alpha = 0.3, width = 0.5) +  
  geom_boxplot(alpha = 0.3, width = 0.1, outliers = FALSE) +  
  scale_fill_manual(values = c("healthy_donors" = "skyblue4", "cdiff_patients" = "indianred4")) +
  geom_point(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.5, seed = 42), 
             aes(group = patient_type, colour = butyrate_levels, shape = butyrate_levels), 
             alpha = 0.75, size = 2.5) +  
  scale_colour_manual(values = c("High butyrate" = "forestgreen", "Low butyrate" = "orange", "Other" = "grey60"), name = NULL) +  
  scale_shape_manual(values = c("High butyrate" = 15, "Low butyrate" = 17, "Other" = 3)) +  
  scale_x_discrete(label = c("Dysbiotic patients", "Healthy donors")) +  
  scale_y_continuous(limits = c(0, 300), breaks = c(0, 100, 200, 300)) +
  labs(x = NULL, y = "Normalized butyrate levels") +  
  guides(colour = guide_legend(order = 2, override.aes = list(shape = c(15, 17, 3))),
         shape = "none", fill = "none") +  
  theme_light() +  
  theme(text = element_text(size = 16),  
        axis.title.y = element_text(vjust = +2, size = 16),  
        legend.position = "bottom",  
        legend.box = "vertical",  # Stack legends vertically  
        legend.text = element_text(size = rel(0.8)))

plot_butyrate

#ggsave("Figures_paper/Fig2A.Boxplot_butyrate.jpeg", width = 6, height = 6, units = "in", dpi = 600)
```

### PCA - Non specific regions (Supp. Fig. S6)

```{r}
#Selected columns non-specific ()
which(colnames(ir_skg_total) == "504") #To check the column of the variable

ir_filtered_neg <- ir_skg_total %>% dplyr::select(1, 2:394, 894:994, 1194:2294, 2784:3490)

#PCA
#The variable 'patient_type' (column = 1) is removed before the PCA analysis
pca_skg_neg_reg <- PCA(ir_filtered_neg, ncp = 5, scale.unit = TRUE, quali.sup = 1)

#OUTPUTS PCA
summary(pca_skg_neg_reg)  #Summary of PCA

#Eigenvalues --> a matrix containing all the eigenvalues, the percentage of variance and the cumulative percentage of variance
pca_skg_neg_reg$eig       

#Coordinates (loadings) --> a list of matrices containing all the results for the active variables (coordinates, correlation between variables and axes, square cosine, contributions)
pca_skg_neg_reg$var$coord 

#Individual scores --> a list of matrices containing all the results for the active individuals (coordinates, square cosine, contributions)
comp_ind <- pca_skg_neg_reg$ind$coord 

#Check how many eigenvalues follow the Kaiser's criterium
sum(pca_skg_neg_reg$eig[, 1] > 1)

#Plotting the PCA results with fviz_pca_ind (part of the factoextra package), uses ggplot2 internally 
pca_sgk_neg_reg <- fviz_pca_ind(pca_skg_neg_reg,  
             label="none", 
             habillage = 1, #Categorical data that separates results, here 'patient_type'
             addEllipses = TRUE, 
             ellipse.level = 0.95,
             legend.title = "Patient Type", 
             title = "NON-SPECIFIC REGIONS - SCFAs",
             axes = c(1, 2), # Display PC1 and PC2
             xlab = paste0("PC1 (", round(pca_skg_neg_reg$eig[1, 2], 2), "%)"), #PC1 
             ylab = paste0("PC2 (", round(pca_skg_neg_reg$eig[2, 2], 2), "%)")) + #PC2
  scale_colour_manual(values = c("Healthy donors" = "gold3", "Dysbiotic patients" = "forestgreen")) + 
  theme(text = element_text(size = 10),  
        axis.title.y = element_text(vjust = +2, size = 11),
        axis.title.x = element_text(vjust = +2, size = 11),
        legend.position = "bottom",  
        legend.box = "vertical",  # Stack legends vertically  
        legend.text = element_text(size = rel(0.8)))

pca_sgk_neg_reg

#ggsave("Figures_paper/SupFigS6.PCA_non_specific_regions.jpeg", width = 6, height = 4, units = "in", dpi = 600)
```

## OTHER SUPPLEMENTARY MATERIALS

### PCA mix samples (Supp. Fig. S5)

```{r}
#Load FTIR data
spc_list_mix <- read_opus(dsn = "~/Library/CloudStorage/OneDrive-UniversitatdeBarcelona/PhD_A.Sayol/2024.IR_Biotyper/FTIR_R_analysis/OPUS_spectra_FTIR/Sample_mix/", data_only = FALSE, parallel = FALSE, progress_bar = FALSE)

#SPECTRA LIST --> Sample mix
ir_names_mix <- names(spc_list_mix)
sample_id_mix <- c("CLA33F", "MPP28F", "RGV45F", "RTF49M", "CLA33F_MPP28F", "RGV45F_RTF49M")

ir_mix <- data.frame()
for (i in seq_along(sample_id_mix)) { #Iterate over each sample to get their 5 replicates
  pattern <- paste0("^", gsub("\\(", "\\\\(", gsub("\\)", "\\\\)", sample_id_mix[i])))
  replicates <- ir_names_mix[grep(pattern, ir_names_mix)]
  if (length(replicates) != 0){
      selected_matrices <- lapply(replicates, function(element) {
      return(t(spc_list_mix[[element]]$ab$data))
  })
  df_with_all_replicates <- as.data.frame(do.call(cbind, selected_matrices))
  df_for_each_sample <- data.frame(row.names = sample_id_mix[i],t((average_spec = rowMeans(df_with_all_replicates))))
  ir_mix <- rbind(ir_mix, df_for_each_sample)
  }
}

#Remove the X before each wavelenght
colnames(ir_mix) <- gsub("^X", "", colnames(ir_mix))

#SPECTRA PROCESSING
#STANDARD NORMAL VARIATE --> With average data (the spectra form does not change, but changes to enter numbers, making it easier to understand)
ir_mix_snv <- data.frame()
for (i in 1:nrow(ir_mix)){
  print(i)
  ir_mix_snv <- rbind(ir_mix_snv,data.frame(standardNormalVariate(ir_mix[i,])))
}
rownames(ir_mix_snv) <- rownames(ir_mix)
ir_mix_snv

#BASELINE CORRECTION --> Applied to SNV corrected data
ir_mix_baseline <- data.frame()
for (i in 1:nrow(ir_mix_snv)){
  numeric_vector <- ir_mix_snv[i,]
  row_df <- data.frame(matrix(baseline(numeric_vector), nrow = 1, ncol = length(numeric_vector)))
  ir_mix_baseline <- rbind(ir_mix_baseline,row_df)
}

colnames(ir_mix_baseline) <- colnames(ir_mix)
rownames(ir_mix_baseline) <- rownames(ir_mix)

#SAVITZKY-GOLAY --> 2nd derivate
ir_mix_sgk <- data.frame()
for (i in 1:nrow(ir_mix_baseline)){
  numeric_vector <- ir_mix_baseline[i,]
  row_df <- data.frame(savitzkyGolay(X = numeric_vector, m = 2, p = 3, w = 9))
  ir_mix_sgk <- rbind(ir_mix_sgk,row_df)
}

#PRINCIPAL COMPONENT ANALYSIS (PCA)
#Merge data from Savitzky-Golay (2nd derivate)
#For the PCA analysis we need a single dataframe with n rows (individuals) and p columns (numeric variables)
ir_mix_pca <- ir_mix_sgk %>% mutate(sample = rep(c("sample_1", "sample_2","sample_3", "sample_4", "mix_1-2", "mix_3-4"))) %>% dplyr::select(sample, everything()) #Places `sample` at the begining

#PCA
#The variable 'sample' (column = 1) is removed before the PCA analysis
pca_mix <- PCA(ir_mix_pca, ncp = 5, scale = TRUE, quali.sup = 1)

#OUTPUTS PCA
summary(pca_mix)  #Summary of PCA

#Eigenvalues --> a matrix containing all the eigenvalues, the percentage of variance and the cumulative percentage of variance
pca_mix$eig       

#Coordinates (loadings) --> a list of matrices containing all the results for the active variables (coordinates, correlation between variables and axes, square cosine, contributions)
pca_mix$var$coord 

#Individual scores --> a list of matrices containing all the results for the active individuals (coordinates, square cosine, contributions)
comp_ind <- pca_mix$ind$coord 

#Plotting the PCA results with fviz_pca_ind (part of the factoextra package), uses ggplot2 internally 
pca_mix_sgk <- fviz_pca_ind(pca_mix,  
             label="none", 
             habillage = 1, #Categorical data that separates results, here 'sample'
             addEllipses = FALSE, 
             ellipse.level = 0.95,
             legend.title = "Sample", 
             axes = c(1, 2), # Display PC1 and PC2
             xlab = paste0("PC1 (", round(pca_mix$eig[1, 2], 2), "%)"), #PC1 
             ylab = paste0("PC2 (", round(pca_mix$eig[2, 2], 2), "%)")) + #PC2
  scale_colour_manual(values = c("sample_1" = "orange2", "sample_2" = "darkorange2", "mix_1-2" = "orangered2", "sample_3" = "olivedrab2", "sample_4" = "darkolivegreen3","mix_3-4" = "chartreuse4"),       
                      breaks = c("sample_1", "sample_3", "sample_2", "sample_4", "mix_1-2", "mix_3-4"), #Define order               
                      labels = c("Sample 1", "Sample 3", "Sample 2", "Sample 4", "Mix 1-2", "Mix 3-4")) + #Custom legend labels       
   scale_shape_manual(values = c("sample_1" = 16, "sample_2" = 16, "mix_1-2" = 1, "sample_3" = 15, "sample_4" = 15, "mix_3-4" = 0), breaks = c("sample_1", "sample_3", "sample_2", "sample_4", "mix_1-2", "mix_3-4"), labels = c("Sample 1", "Sample 3", "Sample 2", "Sample 4", "Mix 1-2", "Mix 3-4")) + #Custom legend labels       
  theme(text = element_text(size = 10),  
        axis.title.y = element_text(vjust = +2, size = 11),
        axis.title.x = element_text(vjust = +2, size = 11),
        legend.position = "bottom",  
        legend.text = element_text(size = rel(0.8)))

pca_mix_sgk

#ggsave("Figures_paper/SupFigS5.PCA_mixed_samples.jpeg", width = 6, height = 4, units = "in", dpi = 600)
```

### Plot acidified H2O (Supp. Fig. S7)

```{r}
#Load FTIR DATA
spc_list_water <- read_opus(dsn = "~/Library/CloudStorage/OneDrive-UniversitatdeBarcelona/PhD_A.Sayol/2024.IR_Biotyper/FTIR_R_analysis/OPUS_spectra_FTIR/Acidified_water/", data_only = FALSE, parallel = FALSE, progress_bar = FALSE)

#SPECTRA LIST --> Acidified water
ir_names_water <- names(spc_list_water)
sample_id_water <- c("Acidified_water", "Acidified_water", "Acidified_water", "Acidified_water", "Acidified_water")

ir_water <- data.frame()
for (i in seq_along(sample_id_water)) { #Iterate over each sample to get their 5 replicates
  pattern <- paste0("^", gsub("\\(", "\\\\(", gsub("\\)", "\\\\)", sample_id_water[i])))
  replicates <- ir_names_water[grep(pattern, ir_names_water)]
  if (length(replicates) != 0){
      selected_matrices <- lapply(replicates, function(element) {
      return(t(spc_list_water[[element]]$ab$data))
  })
  df_with_all_replicates <- as.data.frame(do.call(cbind, selected_matrices))
  df_for_each_sample <- data.frame(row.names = sample_id_water[i],t((average_spec = rowMeans(df_with_all_replicates))))
  ir_water <- rbind(ir_water, df_for_each_sample)
  }
}

#Remove the X before each wavelenght
colnames(ir_water) <- gsub("^X", "", colnames(ir_water))

#SPECTRA PROCESSING
#STANDARD NORMAL VARIATE --> With average data (the spectra form does not change, but changes to enter numbers, making it easier to understand)
ir_water_snv <- data.frame()
for (i in 1:nrow(ir_water)){
  print(i)
  ir_water_snv <- rbind(ir_water_snv,data.frame(standardNormalVariate(ir_water[i,])))
}
rownames(ir_water_snv) <- rownames(ir_water)
ir_water_snv

#BASELINE CORRECTION --> Applied to SNV corrected data
ir_water_bsln <- data.frame()
for (i in 1:nrow(ir_water_snv)){
  numeric_vector <- ir_water_snv[i,]
  row_df <- data.frame(matrix(baseline(numeric_vector), nrow = 1, ncol = length(numeric_vector)))
  ir_water_bsln <- rbind(ir_water_bsln,row_df)
}

colnames(ir_water_bsln) <- colnames(ir_water)
rownames(ir_water_bsln) <- rownames(ir_water)

#PLOT BASELINE CORRECTED DATA
bsln_data_water <- ir_water_bsln %>% 
  rownames_to_column(var = "sample_id") %>%  #Row names to column
  pivot_longer(cols = -sample_id,            #Select all columns except "sample_id"
               names_to = "wavelength",     #Name for the wavelength column
               values_to = "absorbance")    #Name for the absorbance values

#Remove the 'X' before the wavelength value 
bsln_data_water$wavelength <- gsub("X", "", bsln_data_water$wavelength)  
bsln_data_water$wavelength <- as.numeric(bsln_data_water$wavelength)

#PLOT (SNV + baseline correction plot acidified water)
blsn_water_plot <- ggplot(bsln_data_water, aes(x = wavelength, y = absorbance)) + 
  geom_line(linewidth = 0.8, color = "lightblue") + 
  labs(x = expression("Wavenumber (cm"^-1*")"), y = "Absorbance (a.u.)") +
  scale_x_continuous(breaks = seq(min(bsln_data_water$wavelength), max(bsln_data_water$wavelength), by = 200), 
                     labels = seq(min(bsln_data_water$wavelength), max(bsln_data_water$wavelength), by = 200),
                     expand = c(0, 0), trans = "reverse") +
  theme_minimal() +
  theme(text = element_text(size = 9.5),
        axis.title.y = element_text(vjust = +3, size = 11.5), 
        axis.title.x = element_text(vjust = +1, size = 11.5), 
        axis.ticks.x=element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom", 
        legend.text=element_text(size = 9.5))
  
blsn_water_plot

#ggsave("Figures_paper/SupFigS7.Spectra_water.jpeg", width = 6, height = 4, units = "in", dpi = 600)
```

## 
